\textbf{Evaluation of machine learning methods for 16S rRNA gene data}

### 2019-05-24 submission/manuscript.Rmd
update manuscript

>>>>

Members of the human microbiota have been implicated in a wide range of chronic diseases and can be used as biomarkers to diagnose them. However, identifying a set of biomarkers and making generalizable disease diagnoses is challenging because human-associated microbiome is high-dimensional and uneven within and between individuals with the same disease. Machine learning (ML) modeling of human-associated microbial communities is used to overcome this challenge. Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but reliable and reproducible ML models are scarce due to the flawed modeling methods and a lack of transparency in application and selection of ML models. We established a standard for ML modeling pipelines that use 16 rRNA gene sequences to predict diseases. We used fecal 16S rRNA sequences to predict the presence/absence of colorectal cancer (CRC) lesions in patients (N=490) from a previously published CRC study. We developed seven ML models with different classification algorithms (L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and extreme gradient boosted decision tree (XGBoost)) to show the effect of model selection on model accuracy, interpretability and computational efficiency. Random forest model was best at detecting CRC lesions but it was slow to train and hard to interpret. Despite the simplicity, the L2-regularized logistic regression followed random forest in predictive performance and it trained the fastest. Our results in comparing seven ML models with one dataset, showed the importance of model selection in ML model development. It also showcased a robust ML pipeline that creates generalizable and reproducible ML models.

## Importance

### 2019-05-24 submission/manuscript.tex
update manuscript

@@

We evaluated the prediction performances of seven binary classification
models when applied to held-out test data using the area under the
receiver operating characteristic curve (AUROC) as the discriminative
performance metric. Random forest had significantly higher test AUROC
values than the other models for detecting SRNs. The median AUROC of
random forest was 0.695 (IQR 0.044). L2 logistic regression, XGBoost, L2
SVM with linear and radial basis function kernel AUROC values were not
significantly different from one another. They had median AUROC values
of 0.68 (IQR 0.055), 0.679 (IQR 0.053), 0.678 (IQR 0.056) and 0.668 (IQR
0.056) respectively. L1 SVM with linear kernel and decision tree had
significantly lower AUROC values than the other ML models with median
AUROC of 0.65 (IQR 0.066) and 0.601 (IQR 0.059), respectively {[}Figure
2{]}.

>>>>

Members of the human microbiota have been implicated in a wide range of
chronic diseases and can be used as biomarkers to diagnose them.
However, identifying a set of biomarkers and making generalizable
disease diagnoses is challenging because human-associated microbiome is
high-dimensional and uneven within and between individuals with the same
disease. Machine learning (ML) modeling of human-associated microbial
communities is used to overcome this challenge. Progress has been made
towards developing ML models that predict health outcomes from bacterial
abundances, but reliable and reproducible ML models are scarce due to
the flawed modeling methods and a lack of transparency in application
and selection of ML models. We established a standard for ML modeling
pipelines that use 16 rRNA gene sequences to predict diseases. We used
fecal 16S rRNA sequences to predict the presence/absence of colorectal
cancer (CRC) lesions in patients (N=490) from a previously published CRC
study. We developed seven ML models with different classification
algorithms (L2-regularized logistic regression, L1 and L2 support vector
machines (SVM) with linear and radial basis function kernels, a decision
tree, random forest and extreme gradient boosted decision tree
(XGBoost)) to show the effect of model selection on model accuracy,
interpretability and computational efficiency. Random forest model was
best at detecting CRC lesions but it was slow to train and hard to
interpret. Despite the simplicity, the L2-regularized logistic
regression followed random forest in predictive performance and it
trained the fastest. Our results in comparing seven ML models with one
dataset, showed the importance of model selection in ML model
development. It also showcased a robust ML pipeline that creates
generalizable and reproducible ML models.

\subsection{Importance}\label{importance}

@@

We evaluated the prediction performances of seven binary classification
models when applied to held-out test data using the area under the
receiver operating characteristic curve (AUROC) as the discriminative
performance metric {[}Figure 1{]}. Random forest had significantly
higher test AUROC values than the other models for detecting SRNs. The
median AUROC of random forest was 0.695 (IQR 0.044). L2 logistic
regression, XGBoost, L2 SVM with linear and radial basis function kernel
AUROC values were not significantly different from one another. They had
median AUROC values of 0.68 (IQR 0.055), 0.679 (IQR 0.053), 0.678 (IQR
0.056) and 0.668 (IQR 0.056) respectively. L1 SVM with linear kernel and
decision tree had significantly lower AUROC values than the other ML
models with median AUROC of 0.65 (IQR 0.066) and 0.601 (IQR 0.059),
respectively {[}Figure 2{]}.
